---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# semseeker

<!-- badges: start -->
```{r, echo = FALSE, results='asis'}
library("badger")
cat(
  badge_devel("drake69/semseeker", "blue"),
  badge_lifecycle("experimental"),
  badge_codecov("drake69/semseeker"),
  badge_last_commit("drake69/semseeker"),
  badge_github_actions("drake69/semseeker")
)
  # badge_repostatus("Active"),
  # badge_codefactor("drake69/semseeker")
```

The goal of semseeker is to find all methylation localized and enriched variants.

## Installation

You can install semseeker using devtools, future relese will be
available through CRAN.

Install the latest release:

    install.packages("devtools")
    library("devtools")
    install_github("drake69/semseeker")

## Quick Example

This is a basic example which shows how you can create the beta’s
methylation matrix to use for calculation using ChAMP:


    library(ChAMP)
    idat_folder <- "~/source_idat/"
    resultFolder = "~/result/"
    myLoadN <- champ.load(directory = idat_folder,
                          method = "minfi",
                          methValue="B",
                          autoimpute=TRUE,
                          filterDetP=TRUE,
                          ProbeCutoff=0,
                          SampleCutoff=0.1,
                          detPcut=0.01,
                          filterBeads=TRUE,
                          beadCutoff=0.05,
                          filterNoCG=TRUE,
                          filterSNPs=TRUE,
                          population=NULL,
                          filterMultiHit=TRUE,
                          filterXY=TRUE,
                          force=FALSE,
                          arraytype="450K")

    # normalize with ChAMP
    myNormN<-champ.norm(beta=myLoadN$beta,
                        rgSet=myLoadN$rgSet,
                        mset=myLoadN$mset,
                        resultsDir= resultFolder,
                        method="SWAN",
                        plotBMIQ=FALSE,
                        arraytype="450K",
                        cores= detectCores(all.tests = FALSE, logical = TRUE) - 1
                        )

    saveRDS(myNormN,"~/normalizedData.rds")

This how to obtain the analyzed data:

    library(semseeker)

    normalizedData <- readRDS("~/normalizedData.rds")

    sample_sheet <- read.csv2("~/sample_sheet.csv")

    semseeker (sample_sheet = sample_sheet, 
            methylation_data = normalizedData,
            resultFolder = "~/semseeker_result/")

# Complete Example

Look in to the example folder of the repository to seea complete and working example with data from Gene Expression Omnibus (GEO)

# Input requirements
- the samplesheet dataframe should contain a column called Sample_Group, the admitted values are: Case, Control, Reference. if you don0t have the Refernce population you can duplicate the Control population rows and use Reference in the Sample_Group column.
