---
title: "Association analysis ŌĆö statistical models, correlation and group tests"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: association-analysis.bib
vignette: >
  %\VignetteIndexEntry{Association analysis ŌĆö statistical models, correlation and group tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r synthetic_setup, include=FALSE}
# Tiny synthetic sample sheet so chunks below can be evaluated without
# requiring real cohort data on disk.
sample_sheet <- data.frame(
  Sample_ID    = paste0("S", sprintf("%03d", 1:6)),
  Sample_Group = c("Reference","Reference","Reference","Case","Case","Case"),
  age          = c(34, 41, 29, 52, 47, 61),
  bmi          = c(22.1, 25.4, 19.8, 28.6, 26.3, 31.1),
  sex          = c("F","M","F","M","F","M"),
  stringsAsFactors = FALSE
)
```

## Overview

`association_analysis()` is the downstream statistical engine of semseeker.
After `semseeker()` has computed epimutation counts, lesion counts and delta
metrics (`DELTAS`, `DELTAR`, and their ranked variants `DELTAP`/`DELTARP` and
`DELTAQ`/`DELTARQ`), this function tests whether any of those values are
associated with phenotypic or clinical variables.

The function supports **three broad families** of models, all run per genomic
region in parallel using `future` [@future]:

| Family | Methods |
|---|---|
| **Correlation** | Pearson, Spearman, Kendall, permuted Spearman, JensenŌĆōShannon divergence |
| **Group tests** | Wilcoxon (Mann-Whitney), paired Wilcoxon, Student's t-test, paired t-test, Kruskal-Wallis, chi-squared, Fisher's exact, Bartlett, mean permutation |
| **Regression** | Gaussian/Poisson/Binomial GLM, quantile regression (standard and permuted), polynomial, multinomial, nonlinear (logŌéüŌéĆ, powŌéüŌéĆ, log, exp), mediation (quantreg, ridge, linear) |

Multiple comparisons are adjusted with the Benjamini-Hochberg FDR procedure
[@BH1995]. All analyses are run across every combination of genomic area, marker
and sample group defined during the `semseeker()` call.

<hr>

## Prerequisites

You need a completed semseeker run (see the
[Getting started vignette](getting-started.html)).
The sample sheet used in the original `semseeker()` call must already contain
the clinical variable(s) you want to test:

```{r sample_sheet}
# In practice: sample_sheet <- read.csv2("sample_sheet.csv").
# Here we use the synthetic sample sheet defined in the hidden setup chunk.

# Required: Sample_ID, Sample_Group
# Plus any clinical covariates you plan to test:
# age, bmi, treatment, exposure_level, ...
head(sample_sheet)
```

<hr>

## The `inference_details` data frame

Every analysis is specified through an `inference_details` data frame.
Each row is one analysis to run. The columns are:

| Column | Required | Description |
|---|---|---|
| `formula` | Ō£ģ | R-style formula string: `"DELTARP ~ age"` or `"MUTATIONS_HYPO ~ age + sex"` |
| `family` | Ō£ģ | Model/test type (see full table below) |
| `transformation_y` | Ō£ģ | Pre-transformation on the dependent variable |
| `depth_analysis` | Ō£ģ | Granularity of results (1ŌĆō3, see below) |
| `filter_p_value` | Ō£ģ | `TRUE` = keep only FDR-significant results |
| `samples_sql_condition` | Ō¼£ | SQL WHERE fragment to subset samples (e.g. `"Sample_Group = 'Case'"`) |
| `covariates` | Ō¼£ | Comma-separated string of covariate column names |
| `covariates_dummy` | Ō¼£ | Covariates to one-hot encode before modelling |
| `collinearity_check` | Ō¼£ | `TRUE` = remove collinear covariates automatically |
| `covariates_pca` | Ō¼£ | `TRUE` = replace covariates with their first PCs |

<hr>

## Dependent variable: what can be tested

Any semseeker output metric can be the dependent variable in `formula`:

| Metric | Description |
|---|---|
| `MUTATIONS_HYPO` | Count of hypomethylated mutations per region |
| `MUTATIONS_HYPER` | Count of hypermethylated mutations per region |
| `LESIONS_HYPO` | Count of hypomethylated lesions per region |
| `LESIONS_HYPER` | Count of hypermethylated lesions per region |
| `DELTAS` | Delta Signal ŌĆö signed deviation of the value beyond the reference threshold |
| `DELTAR` | Delta Ratio ŌĆö `DELTAS` divided by the reference interval width (superior ŌłÆ inferior) |
| `DELTAP` | `DELTAS` discretised into equal-width bins, each assigned an integer rank weight |
| `DELTARP` | `DELTAR` discretised into equal-width bins, each assigned an integer rank weight (default 4 bins) |
| `DELTAQ` | `DELTAS` discretised into quantile bins, each assigned an integer rank weight |
| `DELTARQ` | `DELTAR` discretised into quantile bins, each assigned an integer rank weight (default 4 bins) |
| `SIGNAL_MEAN` | Per-probe mean methylation signal (marker `SIGNAL`, figure `MEAN`) |

<hr>

## All supported model types (`family`)

### Correlation

```{r correlation_examples}
# Spearman rank correlation (recommended for non-normal SEM metrics)
inf <- data.frame(formula="DELTARP ~ age", family="spearman",
                  transformation_y="none", depth_analysis=3,
                  filter_p_value=TRUE)

# Pearson linear correlation
inf <- data.frame(formula="DELTARP ~ bmi", family="pearson",
                  transformation_y="log", depth_analysis=3,
                  filter_p_value=FALSE)

# Kendall's tau (more robust with tied ranks)
inf <- data.frame(formula="MUTATIONS_HYPO ~ age", family="kendall",
                  transformation_y="none", depth_analysis=2,
                  filter_p_value=TRUE)

# Spearman with permutation-based p-values (more accurate for small N)
# Syntax: "spearman-permutation_<n_permutations>"
inf <- data.frame(formula="DELTARP ~ age", family="spearman-permutation_1000",
                  transformation_y="none", depth_analysis=3,
                  filter_p_value=TRUE)

# Jensen-Shannon divergence (compares distributions)
inf <- data.frame(formula="DELTARP ~ Sample_Group", family="jsd",
                  transformation_y="none", depth_analysis=2,
                  filter_p_value=TRUE)

# Run the analysis (uncomment when you have a real semseeker results folder):
# association_analysis(inf, result_folder = "~/semseeker_results/")
```

Spearman [@spearman1904] is the default recommendation for SEM metrics because
they are often skewed and non-normally distributed. Pearson assumes bivariate
normality. Kendall's tau [@kendall1938] is preferable when there are many tied
values.

<hr>

### Group comparison tests

```{r group_tests}
# Wilcoxon rank-sum / Mann-Whitney (two independent groups)
inf <- data.frame(formula="MUTATIONS_HYPO ~ Sample_Group", family="wilcoxon",
                  transformation_y="none", depth_analysis=2,
                  filter_p_value=TRUE)

# Paired Wilcoxon (repeated measures / matched samples)
inf <- data.frame(formula="DELTARP ~ timepoint", family="wilcoxon.paired_timepoint",
                  transformation_y="none", depth_analysis=2,
                  filter_p_value=TRUE)

# Student's t-test
inf <- data.frame(formula="DELTARP ~ Sample_Group", family="t.test",
                  transformation_y="scale", depth_analysis=3,
                  filter_p_value=TRUE)

# Paired t-test
inf <- data.frame(formula="DELTARP ~ timepoint", family="t.test.paired_timepoint",
                  transformation_y="none", depth_analysis=3,
                  filter_p_value=TRUE)

# Kruskal-Wallis (> 2 groups, non-parametric ANOVA)
inf <- data.frame(formula="MUTATIONS_HYPO ~ disease_stage", family="kruskal.test",
                  transformation_y="none", depth_analysis=2,
                  filter_p_value=TRUE)

# Chi-squared test (categorical dependent variable)
inf <- data.frame(formula="MUTATIONS_HYPO ~ exposure", family="chisq.test",
                  transformation_y="none", depth_analysis=1,
                  filter_p_value=TRUE)

# Fisher's exact test (small samples, 2├Ś2 tables)
inf <- data.frame(formula="MUTATIONS_HYPO ~ binary_exposure", family="fisher.test",
                  transformation_y="none", depth_analysis=1,
                  filter_p_value=TRUE)

# Bartlett's test for equality of variances
inf <- data.frame(formula="DELTARP ~ Sample_Group", family="bartlett.test",
                  transformation_y="none", depth_analysis=2,
                  filter_p_value=TRUE)

# Mean permutation test (non-parametric, permutation-based)
# Syntax: "mean-permutation_<n_permutations>"
inf <- data.frame(formula="DELTARP ~ Sample_Group", family="mean-permutation_1000",
                  transformation_y="none", depth_analysis=3,
                  filter_p_value=TRUE)

# Uncomment when you have a real semseeker results folder:
# association_analysis(inf, result_folder = "~/semseeker_results/")
```

The Wilcoxon rank-sum test [@wilcoxon1945] is the recommended non-parametric
alternative to the t-test when the normality assumption cannot be verified, which
is common for mutation count data.

<hr>

### Regression models (GLM)

```{r glm}
# Gaussian linear regression (continuous outcome)
# Formula can include multiple predictors
inf <- data.frame(formula="DELTARP ~ age + sex + bmi",
                  family="gaussian",
                  transformation_y="log",
                  depth_analysis=3,
                  filter_p_value=TRUE,
                  covariates="sex,bmi")   # declare covariates explicitly

# Poisson regression (count data ŌĆö mutations, lesions)
inf <- data.frame(formula="MUTATIONS_HYPO ~ exposure_level",
                  family="poisson",
                  transformation_y="none",
                  depth_analysis=2,
                  filter_p_value=TRUE)

# Binomial / logistic regression (binary outcome)
inf <- data.frame(formula="case_control ~ DELTARP",
                  family="binomial",
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Multinomial regression (> 2 categories)
inf <- data.frame(formula="disease_stage ~ DELTARP",
                  family="multinomial",
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Uncomment when you have a real semseeker results folder:
# association_analysis(inf, result_folder = "~/semseeker_results/")
```

GLMs [@glm-nelder] are appropriate when the relationship between the SEM metric
and the clinical variable is expected to be linear (Gaussian), when the
outcome is a count (Poisson), or when predicting a binary endpoint from SEM
scores (Binomial).

<hr>

### Quantile regression

```{r quantreg}
# Standard quantile regression at quantile Žä
# Syntax: "quantreg_<tau>"
inf <- data.frame(formula="DELTARP ~ age",
                  family="quantreg_0.5",       # median regression
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Quantile regression with permutation-based inference
# Syntax: "quantreg-permutation_<tau>_<n_permutations>"
inf <- data.frame(formula="DELTARP ~ age",
                  family="quantreg-permutation_0.5_1000",
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Quantile permutation model
# Syntax: "quantile-permutation_<n_permutations>"
inf <- data.frame(formula="DELTARP ~ Sample_Group",
                  family="quantile-permutation_1000",
                  transformation_y="none",
                  depth_analysis=2,
                  filter_p_value=TRUE)

# Uncomment when you have a real semseeker results folder:
# association_analysis(inf, result_folder = "~/semseeker_results/")
```

Quantile regression [@quantreg-koenker; @lqmm] is particularly useful for SEM
metrics: unlike mean regression it is robust to outliers and captures the full
conditional distribution, which matters when only high-quantile epigenetic
variability is biologically relevant.

<hr>

### Nonlinear regression (NLS)

```{r nls}
# LogŌéüŌéĆ dose-response model
# Syntax: "log10_<scale_factor>"
inf <- data.frame(formula="DELTARP ~ dose",
                  family="log10_1",
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Power of 10 model
inf <- data.frame(formula="DELTARP ~ dose",
                  family="pow10_1",
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Natural log model
inf <- data.frame(formula="DELTARP ~ concentration",
                  family="log_1",
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Uncomment when you have a real semseeker results folder:
# association_analysis(inf, result_folder = "~/semseeker_results/")
```

<hr>

### Polynomial regression

```{r polynomial}
# Syntax: "polynomial_<degree>"
inf <- data.frame(formula="DELTARP ~ age",
                  family="polynomial_2",       # quadratic
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Uncomment when you have a real semseeker results folder:
# association_analysis(inf, result_folder = "~/semseeker_results/")
```

<hr>

### Mediation analysis

Mediation models test whether a third variable (mediator) explains the
relationship between the SEM metric and an outcome:

```{r mediation}
# Mediation via quantile regression
inf <- data.frame(formula="DELTARP ~ exposure + mediator",
                  family="mediation-quantreg_0.5",
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Mediation via ridge regression (regularised, handles multicollinearity)
inf <- data.frame(formula="DELTARP ~ exposure + mediator",
                  family="mediation-ridge",
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Mediation via linear model
inf <- data.frame(formula="DELTARP ~ exposure + mediator",
                  family="mediation-linear",
                  transformation_y="none",
                  depth_analysis=3,
                  filter_p_value=TRUE)

# Uncomment when you have a real semseeker results folder:
# association_analysis(inf, result_folder = "~/semseeker_results/")
```

<hr>

## Covariates and confounders

Include covariates directly in the formula and declare them in the `covariates`
column so semseeker correctly separates the independent variable of interest
from confounders:

```{r covariates}
inf <- data.frame(
  formula          = "DELTARP ~ tcdd_exposure + age + sex + bmi",
  family           = "gaussian",
  transformation_y = "log",
  depth_analysis   = 3,
  filter_p_value   = TRUE,
  covariates       = "age,sex,bmi",        # confounders to adjust for
  collinearity_check = TRUE,               # auto-remove collinear covariates
  covariates_pca   = FALSE                 # TRUE = replace with PCs
)

# Uncomment when you have a real semseeker results folder:
# association_analysis(inf, result_folder = "~/semseeker_results/")
```

When `collinearity_check = TRUE`, semseeker automatically detects and removes
collinear covariates using VIF before fitting the model. When `covariates_pca = TRUE`,
covariates are replaced by their principal components before fitting.

<hr>

## Filtering by sample subset (`samples_sql_condition`)

Run the analysis on a subset of samples using a SQL WHERE clause:

```{r sql_filter}
# Only Case samples
inf <- data.frame(
  formula               = "DELTARP ~ age",
  family                = "spearman",
  transformation_y      = "none",
  depth_analysis        = 3,
  filter_p_value        = TRUE,
  samples_sql_condition = "Sample_Group = 'Case'"
)

# Only samples with age > 40
inf <- data.frame(
  formula               = "DELTARP ~ bmi",
  family                = "pearson",
  transformation_y      = "none",
  depth_analysis        = 3,
  filter_p_value        = TRUE,
  samples_sql_condition = "age > 40"
)

# Uncomment when you have a real semseeker results folder:
# association_analysis(inf, result_folder = "~/semseeker_results/")
```

<hr>

## Running multiple analyses at once

Combine multiple rows in `inference_details` to run all analyses in a single call:

```{r multi}
inference_details <- rbind(
  data.frame(formula="DELTARP ~ age",             family="spearman",
             transformation_y="none",             depth_analysis=3,
             filter_p_value=TRUE,                 covariates=NA_character_,
             stringsAsFactors=FALSE),
  data.frame(formula="MUTATIONS_HYPO ~ Sample_Group", family="wilcoxon",
             transformation_y="none",             depth_analysis=2,
             filter_p_value=TRUE,                 covariates=NA_character_,
             stringsAsFactors=FALSE),
  data.frame(formula="DELTARP ~ bmi",             family="quantreg_0.5",
             transformation_y="log",              depth_analysis=3,
             filter_p_value=TRUE,                 covariates=NA_character_,
             stringsAsFactors=FALSE),
  data.frame(formula="MUTATIONS_HYPO ~ exposure + age + sex", family="poisson",
             transformation_y="none",             depth_analysis=3,
             filter_p_value=TRUE,                 covariates="age,sex",
             stringsAsFactors=FALSE)
)

# Uncomment when you have a real semseeker results folder:
# association_analysis(
#   inference_details = inference_details,
#   result_folder     = "~/semseeker_results/",
#   parallel_strategy = "multicore",
#   maxResources      = 90
# )
```

<hr>

## Transformation options (`transformation_y`)

Pre-transformations are applied to the SEM metric before testing:

| Value | Transformation | When to use |
|---|---|---|
| `"none"` | No transformation | Metric already approximately normal |
| `"log"` | Natural log: ln(y + ╬Ą) | Right-skewed counts |
| `"log2"` | LogŌéé | Fold-change interpretation |
| `"log10"` | LogŌéüŌéĆ | Wide dynamic range |
| `"scale"` | Z-score: (y ŌłÆ ╬╝) / Žā | Comparing coefficients across regions |
| `"exp"` | Exponential: e^y | Log-scale input |
| `"quantile_3"` | Quantile normalisation, 3 bins | Non-parametric normalisation |

<hr>

## Depth levels (`depth_analysis`)

| Value | Breakdown | Use case |
|---|---|---|
| `1` | Sample-level aggregate only | Quick per-sample overview |
| `2` | Per marker type (MUTATIONS, LESIONS, DELTARP, ŌĆ”) | Identify which metric drives the association |
| `3` | Per genomic area (GENE body, TSS200, TSS1500, island, ŌĆ”) | Fine-grained genomic localisation |

Higher depth levels include all lower levels.

<hr>

## Output files

Results are written to `result_folder/Inference/`. File names encode the full
analysis specification:

```
Inference/
Ōö£ŌöĆŌöĆ DELTARP_GENE_spearman_age_depth3.csv
Ōö£ŌöĆŌöĆ MUTATIONS_HYPO_GENE_wilcoxon_Sample_Group_depth2.csv
Ōö£ŌöĆŌöĆ DELTARP_GENE_gaussian_tcdd_exposure_depth3.csv
ŌööŌöĆŌöĆ ...
```

Each CSV contains:

| Column | Description |
|---|---|
| `CHR`, `START`, `END` | Genomic coordinates |
| `AREA`, `SUBAREA` | Aggregation level and sub-region |
| `MARKER` | SEM metric tested |
| `SCORE` | Test statistic (Žü, W, t, ╬▓, ŌĆ”) |
| `P_Value` | Nominal p-value |
| `Q` / `PVALUE_ADJ_ALL_BH` | FDR-adjusted p-value (Benjamini-Hochberg) |
| `SIGNIFICATIVE_ADJ` | `TRUE` if Q < ╬▒ (default 0.05) |
| `FAMILY_TEST` | Model used |
| `SAMPLE_GROUP` | Sample subset tested |

<hr>

## Reading and exploring results

```{r read_results, eval=FALSE}
# Read a Spearman result file
results <- read.csv(
  list.files("~/semseeker_results/Inference",
             pattern = "spearman.*age.*depth3\\.csv$",
             full.names = TRUE)[1]
)

# Top significant genes by adjusted p-value
sig <- subset(results, SIGNIFICATIVE_ADJ == TRUE)
head(sig[order(sig$Q), c("CHR","START","END","AREA","SCORE","P_Value","Q")], 20)

# Volcano-style: effect size vs -log10(Q)
plot(-log10(results$Q), results$SCORE,
     xlab = "-log10(Q)", ylab = "Spearman rho",
     col  = ifelse(results$SIGNIFICATIVE_ADJ, "red", "grey"),
     pch  = 16, main = "Spearman rho vs FDR")
abline(h = 0, lty = 2)
```

<hr>

## Cross-study and cross-subsample analyses

After running `association_analysis()` on two or more independent cohorts stored
in separate `result_folder`s, semseeker provides functions to compare and
meta-analyse results across studies:

```{r cross_study, eval=FALSE}
# Overlap significant hits across two studies
assoc_inter_study_association_overlaps(
  studies       = list("~/study_A/", "~/study_B/"),
  result_folder = "~/meta_results/",
  pvalue_column = "PVALUE_ADJ_ALL_BH"
)

# Meta-analysis (combines p-values / effect sizes)
assoc_inter_study_association_meta_analysis(
  studies       = list("~/study_A/", "~/study_B/"),
  result_folder = "~/meta_results/"
)

# Subsample stability analysis (bootstrap cohort sub-sampling)
assoc_intra_study_association_subsamples_overlaps(
  result_folder = "~/semseeker_results/",
  n_subsamples  = 100,
  fraction      = 0.8
)
```

<hr>

## Complementary: Bayesian posterior probability

While `association_analysis()` reports frequentist p-values / q-values for each
region, `assoc_bayes_analysis()` provides the *probabilistic complement*: per-region
posterior probabilities computed via empirical Bayes.

For each `(MARKER, AREA, SUBAREA, AREA_OF_TEST)` it estimates:

- `P(case  | epimutated)` ŌĆö how likely a sample is a case given the SEM hit
- `P(control | not-epimutated)` ŌĆö how likely a sample is a control given absence of the SEM hit

A region is reported as a candidate biomarker when both posteriors exceed the
configured thresholds (`bayes_case_threshold` and `bayes_control_threshold`).
Output is written as CSV files under `<result_folder>/Euristic/`, sorted by
posterior probability.

```{r assoc_bayes_analysis, eval=FALSE}
assoc_bayes_analysis(
  result_folder           = "~/semseeker_results/",
  independent_variable    = "Sample_Group",
  bayes_case_threshold    = 0.9,   # min P(case | epimutated)
  bayes_control_threshold = 0.1,   # max P(control | epimutated)
  parallel_strategy       = "multicore",
  maxResources            = 90
)
```

This is most useful when the frequentist q-value ranking is similar across many
regions and you want a probability-based criterion to triage candidates. The two
analyses are independent ŌĆö running both gives you a frequentist and a Bayesian
view of the same data.

<hr>

## Diagnostic performance: sensitivity and specificity

For binary outcomes (case vs control), `diagnostic_performance()` computes per-region
sensitivity, specificity, Jensen-Shannon distance, and a combined diagnostic score ŌĆö
treating each SEM mutation as a candidate diagnostic marker.

```{r diagnostic_performance, eval=FALSE}
diagnostic_performance(
  combinations         = list(c("CASE_A", "CASE_B")),
  result_folder        = "~/semseeker_results/",
  independent_variable = "Sample_Group",
  parallel_strategy    = "multicore"
)
```

Output (per marker) is sorted descending by `SCORE`; the top rows are the
regions with the most promising diagnostic profile in the studied cohort. CSV
files are written to `<result_folder>/Euristic/`.

<hr>

## Cross-study replication

`assoc_intra_study_association_replication()` reruns the association on a target study restricted
to the regions that were already significant in a reference study, then merges
the two result sets. It's the standard "replication / validation" workflow when
you have a discovery cohort and a validation cohort.

```{r cross_study_replication, eval=FALSE}
assoc_intra_study_association_replication(
  inference_details_origin = inference_discovery,
  inference_details        = inference_validation,
  result_folder            = "~/semseeker_replication/",
  parallel_strategy        = "multicore"
)
```

<hr>

## Downstream: pathway and enrichment analysis

Significant genomic regions from `association_analysis()` can be fed directly
into pathway enrichment tools:

```{r pathway_next, eval=FALSE}
# WebGestaltR ŌĆö ORA or GSEA
enrich_WebGestalt(
  study            = "my_study",
  types            = c("BP", "MF"),
  enrich_methods   = c("ORA"),
  inference_detail = inf,
  significance     = TRUE
)

# STRINGdb ŌĆö protein network analysis
enrich_STRINGdb(
  study            = "my_study",
  inference_detail = inf,
  significance     = TRUE
)

# pathfindR ŌĆö active subnetwork enrichment
enrich_pathfindR(
  study            = "my_study",
  inference_detail = inf
)
```

See the [Pathway and enrichment analysis vignette](pathway-analysis.html) for
full details.

<hr>

## R packages used

Statistical models are implemented using R [@R-base] base `stats` functions,
with parallelisation via `future` [@future] and `doFuture`. Quantile regression
uses `lqmm` [@lqmm] and the `quantreg` package [@quantreg-koenker].

## Session info

```{r session_info}
sessionInfo()
```

<hr>

## References
