---
title: "Association analysis — correlation and Wilcoxon test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Association analysis — correlation and Wilcoxon test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

Once semseeker has produced delta metrics (ΔRP, ΔBeta, …) and mutation counts,
you can test whether those values are associated with clinical or phenotypic
variables using `association_analysis()`.

The function supports:

- **Correlation tests**: Pearson, Spearman, Kendall
- **Group comparison tests**: Wilcoxon rank-sum, Student's t-test
- **Regression models**: Gaussian, Poisson, Binomial, Quantile regression

All tests are run per genomic region in parallel and results are adjusted for
multiple comparisons (FDR by default).

---

## Prerequisites

You need a completed semseeker run (see the
[Getting started vignette](getting-started.html)) and a sample sheet with the
clinical variable(s) you want to test.

```{r sample_sheet_with_clin}
# Sample sheet must contain Sample_ID, Sample_Group AND the clinical variable
sample_sheet <- read.csv2("sample_sheet.csv")

# Example columns: Sample_ID, Sample_Group, age, bmi, case_control_binary
head(sample_sheet)
```

---

## Example 1 — Spearman correlation with a continuous variable

Test whether ΔRP per gene correlates with patient age:

```{r spearman}
library(semseeker)

inference_details <- data.frame(
  formula          = "DELTARP ~ age",         # dependent ~ independent
  family           = "spearman",              # correlation type
  transformation_y = "none",                  # no transformation of ΔRP
  depth            = 3,                       # 3 = gene-level breakdown
  filter_p_value   = TRUE                     # keep only significant results
)

association_analysis(
  inference_details = inference_details,
  result_folder     = "~/semseeker_results/",
  parallel_strategy = "multicore",
  maxResources      = 80
)
```

Results are written to `result_folder/Inference/`. The CSV file contains:

| Column | Description |
|---|---|
| `CHR`, `START`, `END` | Genomic coordinates |
| `AREA` | Aggregation level (GENE, ISLAND, …) |
| `rho` | Spearman correlation coefficient |
| `P_Value` | Nominal p-value |
| `Q` | FDR-adjusted p-value |
| `SIGNIFICATIVE_ADJ` | `TRUE` if Q < 0.05 |

```{r read_spearman}
library(polars)

results <- read.csv(
  list.files("~/semseeker_results/Inference", pattern = "spearman.*\\.csv$",
             full.names = TRUE)[1]
)

# Top correlated genes
head(results[order(results$Q), c("AREA", "rho", "P_Value", "Q")], 10)
```

---

## Example 2 — Wilcoxon test (Case vs Control)

Test whether mutation burden (MUTATIONS_HYPO) differs between cases and
controls at gene level:

```{r wilcoxon}
inference_details <- data.frame(
  formula          = "MUTATIONS_HYPO ~ Sample_Group",
  family           = "wilcoxon",
  transformation_y = "none",
  depth            = 2,
  filter_p_value   = TRUE
)

association_analysis(
  inference_details = inference_details,
  result_folder     = "~/semseeker_results/",
  parallel_strategy = "multicore"
)
```

> **Note:** `Sample_Group` must be a binary factor (`"Case"` / `"Control"`) in
> the sample sheet passed to the original `semseeker()` call.

The output CSV adds:

| Column | Description |
|---|---|
| `W` | Wilcoxon W statistic |
| `P_Value` | Nominal p-value (two-sided) |
| `Q` | BH-adjusted p-value |
| `SIGNIFICATIVE_ADJ` | `TRUE` if Q < 0.05 |

---

## Example 3 — Pearson correlation with ΔRP

```{r pearson}
inference_details <- data.frame(
  formula          = "DELTARP ~ bmi",
  family           = "pearson",
  transformation_y = "log",          # log-transform ΔRP before testing
  depth            = 3,
  filter_p_value   = FALSE           # keep all regions
)

association_analysis(
  inference_details = inference_details,
  result_folder     = "~/semseeker_results/"
)
```

---

## Example 4 — Multiple tests in a single call

`inference_details` can be a list of data frames or a multi-row data frame to
run several analyses in one go:

```{r multiple}
inference_details <- rbind(
  data.frame(formula="DELTARP ~ age",          family="spearman",
             transformation_y="none", depth=3, filter_p_value=TRUE),
  data.frame(formula="MUTATIONS_HYPO ~ Sample_Group", family="wilcoxon",
             transformation_y="none", depth=2, filter_p_value=TRUE),
  data.frame(formula="LESIONS_HYPO ~ bmi",     family="pearson",
             transformation_y="log",  depth=3, filter_p_value=TRUE)
)

association_analysis(
  inference_details = inference_details,
  result_folder     = "~/semseeker_results/",
  parallel_strategy = "multicore",
  maxResources      = 90
)
```

---

## Supported families / test types

| `family` value | Test | Dependent variable type |
|---|---|---|
| `"pearson"` | Pearson correlation | Continuous |
| `"spearman"` | Spearman rank correlation | Continuous / ordinal |
| `"kendall"` | Kendall's tau | Continuous / ordinal |
| `"wilcoxon"` | Wilcoxon rank-sum (Mann-Whitney) | Continuous, groups = 2 |
| `"stats::t.test"` | Student's t-test | Continuous, groups = 2 |
| `"gaussian"` | Linear regression (GLM) | Continuous |
| `"poisson"` | Poisson regression | Count data |
| `"binomial"` | Logistic regression | Binary outcome |
| `"quantreg_0.5_1000"` | Quantile regression (τ=0.5, 1000 perms) | Continuous |

---

## Depth levels

The `depth` parameter controls how granularly results are broken down:

| `depth` | Output includes |
|---|---|
| `1` | Per-sample aggregate only |
| `2` | Per marker type (MUTATIONS, LESIONS, DELTARP, …) |
| `3` | Per genomic area (GENE body, TSS200, TSS1500, …) |

---

## Transformation options

Use `transformation_y` to pre-transform the SEM metric before testing:

| Value | Transformation |
|---|---|
| `"none"` | No transformation |
| `"log"` | Natural log |
| `"log2"` | Log base 2 |
| `"log10"` | Log base 10 |
| `"scale"` | Z-score standardisation |
| `"exp"` | Exponential |
| `"quantile_3"` | Quantile normalisation with 3 bins |

---

## Output location

All results land in `result_folder/Inference/` with filenames encoding the
test parameters, e.g.:

```
Inference/
├── DELTARP_GENE_spearman_age_depth3.csv
├── MUTATIONS_HYPO_GENE_wilcoxon_Sample_Group_depth2.csv
└── ...
```

Each file is a CSV sorted by adjusted p-value (`Q`) with only significant
hits when `filter_p_value = TRUE`.
