---
title: "Getting started — epimutations and delta metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started — epimutations and delta metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # code is illustrative; not run during R CMD check
)
```

## What is semseeker?

**semseeker** is an R package for detecting **Stochastic Epigenetic Mutations
(SEMs)** — aberrant, sample-specific changes in DNA methylation that deviate
from the expected population signal.

Starting from a normalised beta-value (or M-value) matrix, semseeker:

1. Identifies **hypomethylated** and **hypermethylated** mutations at
   single-probe resolution.
2. Aggregates adjacent mutations into **lesions** (genomic clusters of
   co-occurring mutations).
3. Computes population-level **delta metrics** (ΔBeta, ΔIV, ΔEV, ΔRV, ΔRP)
   that summarise stochastic epigenetic variability per genomic region.

Supported input platforms: Illumina EPIC (850k), 450k, and 27k arrays
(detected automatically from probe IDs).

---

## Installation

```{r install}
# Install remotes if needed
install.packages("remotes")

# Install semseeker from GitHub
remotes::install_github("drake69/semseeker")
```

> **Note:** semseeker requires the `polars` package from R-multiverse.
> It is installed automatically as a dependency. If you get an installation
> error, try:
> ```r
> Sys.setenv(NOT_CRAN = "true")
> install.packages("polars", repos = "https://community.r-multiverse.org")
> ```

---

## Step 1 — Prepare your data

### Beta-value matrix

semseeker expects a **probe × sample** matrix of beta values (0–1) or M-values.
Probe IDs must be Illumina CpG identifiers (e.g. `cg00000029`).

```{r beta_matrix}
# Example: load a pre-normalised matrix (rows = probes, columns = samples)
beta_matrix <- readRDS("normalised_beta.rds")

dim(beta_matrix)       # e.g. 866895 × 50
beta_matrix[1:3, 1:3]
```

If you are starting from raw IDAT files, normalise first with
[ChAMP](https://bioconductor.org/packages/ChAMP/):

```{r chamap_normalise}
library(ChAMP)

myLoad <- champ.load(
  directory  = "path/to/idat/",
  arraytype  = "EPIC"
)

myNorm <- champ.norm(
  beta      = myLoad$beta,
  rgSet     = myLoad$rgSet,
  mset      = myLoad$mset,
  method    = "SWAN",
  arraytype = "EPIC"
)

beta_matrix <- myNorm
```

### Sample sheet

A data frame with at minimum:

| Column | Description |
|---|---|
| `Sample_ID` | Unique sample identifier — must match column names of the beta matrix |
| `Sample_Name` | Human-readable label |
| `Sample_Group` | One of `"Case"`, `"Control"`, `"Reference"` |

```{r sample_sheet}
sample_sheet <- read.csv2("sample_sheet.csv")

# Minimal required columns
head(sample_sheet[, c("Sample_ID", "Sample_Name", "Sample_Group")])
```

> **Tip:** If you have no independent reference population, duplicate the
> Control rows and set `Sample_Group = "Reference"` for the duplicates.

---

## Step 2 — Detect epimutations

The main entry point is `semseeker()`. A single call runs the full pipeline:
normalisation → mutation calling → lesion aggregation → delta metric computation.

```{r run_semseeker}
library(semseeker)

semseeker(
  sample_sheet  = sample_sheet,
  signal_data   = beta_matrix,
  result_folder = "~/semseeker_results/"
)
```

### Key parameters

| Parameter | Default | Description |
|---|---|---|
| `parallel_strategy` | `"multicore"` | Parallelisation backend (`"sequential"`, `"multicore"`, `"multisession"`) |
| `maxResources` | `90` | % of CPU cores to use |
| `iqrTimes` | `3` | IQR multiplier for aberration threshold |
| `alpha` | `0.05` | Significance threshold |
| `areas` | `c("GENE","CHR","ISLAND","PROBE")` | Genomic aggregation levels |
| `marker` | `c("MUTATIONS","LESIONS","DELTARQ",...)` | Which markers to compute |
| `sex_chromosome_remove` | `TRUE` | Remove sex chromosomes from analysis |

Example — run only probe-level mutations, sequential mode, 2 cores:

```{r run_minimal}
semseeker(
  sample_sheet      = sample_sheet,
  signal_data       = beta_matrix,
  result_folder     = "~/semseeker_results/",
  parallel_strategy = "sequential",
  areas             = c("PROBE"),
  marker            = c("MUTATIONS", "LESIONS")
)
```

---

## Step 3 — Understand the output

Results are written under `result_folder/Data/`. Each sample produces:

```
Data/
├── SAMPLEID_MUTATIONS_HYPO.bed       ← probe-level hypomethylated mutations
├── SAMPLEID_MUTATIONS_HYPER.bed      ← probe-level hypermethylated mutations
├── SAMPLEID_LESIONS_HYPO.bed         ← genomic lesion clusters (hypo)
├── SAMPLEID_LESIONS_HYPER.bed        ← genomic lesion clusters (hyper)
├── SAMPLEID_SIGNAL_MEAN.PROBE.bedgraph
└── ...
```

Population-level pivot tables aggregate all samples per marker:

```
Data/
├── SIGNAL_MEAN_PROBE_WHOLE.parquet   ← mean signal per probe, all samples
├── MUTATIONS_HYPO_PROBE_WHOLE.parquet
├── DELTARP_GENE_WHOLE.parquet        ← ΔRP per gene, all samples
└── ...
```

### Reading results

```{r read_results}
library(polars)

# Read the DELTARP pivot table
deltarp <- as.data.frame(
  pl$read_parquet("~/semseeker_results/Data/DELTARP_GENE_WHOLE.parquet")
)

head(deltarp[, 1:6])
```

---

## Step 4 — Delta metrics explained

semseeker computes five delta metrics per genomic region, each capturing a
different aspect of epigenetic variability:

| Metric | Full name | Captures |
|---|---|---|
| **ΔBeta** | Delta Beta | Mean methylation shift vs reference |
| **ΔIV** | Delta Intra-sample Variability | Within-sample probe dispersion |
| **ΔEV** | Delta Extra-sample Variability | Between-sample variance |
| **ΔRV** | Delta Relative Variability | Normalised inter-sample spread |
| **ΔRP** | Delta Ranked Product | Combined ranked signal — most sensitive |

**ΔRP** (Delta Ranked Product) is the recommended metric for downstream
analysis: it combines mutation burden, lesion frequency and signal variability
into a single ranked score per region, making it robust to outliers and
suitable for association testing.

```{r deltarp_example}
# Load ΔRP at gene level
deltarp_gene <- as.data.frame(
  pl$read_parquet("~/semseeker_results/Data/DELTARP_GENE_WHOLE.parquet")
)

# Top regions by mean ΔRP across cases
case_cols  <- grep("^CASE", names(deltarp_gene), value = TRUE)
deltarp_gene$mean_deltarp <- rowMeans(deltarp_gene[, case_cols], na.rm = TRUE)

head(deltarp_gene[order(-deltarp_gene$mean_deltarp),
                  c("CHR", "START", "END", "mean_deltarp")], 10)
```

---

## Next steps

- **Association analysis** (correlation, Wilcoxon test): see the
  [Association analysis vignette](association-analysis.html).
- **Full function reference**: see the [Reference](../reference/index.html) page.
