---
title: "Getting started — epimutations and delta metrics"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: getting-started.bib
vignette: >
  %\VignetteIndexEntry{Getting started — epimutations and delta metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>"
)
```

```{r real_data_setup, include=FALSE}
# ── Demo dataset: real GSE133774 (EPIC 850k, BWS + MLID) ─────────────────
#
# test_signal_gse133774 and test_samplesheet_gse133774 are bundled fixtures
# built from the GEO series matrix via data-raw/build_test_signal_fixture.R.
# They contain real EPIC beta values for 18k probes around 81 imprinting DMRs
# (10 samples: 6 controls + 4 BWS/MLID family members, including L1 = BWS).
#
# Filtering to the KCNQ1OT1 + H19_IGF2 DMR probes keeps the vignette fast
# while preserving the biologically relevant signal (real BWS epimutations).
data("test_signal_gse133774",      package = "SEMseeker")
data("test_samplesheet_gse133774", package = "SEMseeker")
data("test_master_features",       package = "SEMseeker")

.demo_probes <- subset(test_master_features,
                       DMR_LABEL %in% c("KCNQ1OT1:TSS-DMR", "H19_IGF2:IG-DMR") &
                       PROBE %in% rownames(test_signal_gse133774))
beta_matrix  <- test_signal_gse133774[.demo_probes$PROBE, , drop = FALSE]
sample_sheet <- test_samplesheet_gse133774[, c("Sample_ID", "Sample_Group")]
sample_sheet$Sample_Name <- sample_sheet$Sample_ID
sample_sheet <- sample_sheet[, c("Sample_ID", "Sample_Name", "Sample_Group")]

tmp_result_folder <- file.path(tempdir(), "semseeker_vignette_results")
```

## What is semseeker?

**semseeker** is an R package for detecting, quantifying and analysing
**Stochastic Epigenetic Mutations (SEMs)** from Illumina array-based DNA
methylation data [@Corsaro2025thesis].

### Background: what is a SEM?

A **Stochastic Epigenetic Mutation** is a rare, sample-specific aberration in
DNA methylation at a single CpG site — a probe that deviates from the
population-level signal in an individual sample beyond what can be explained by
normal biological or technical variability [@Gentilini2015].

SEMs were first formally defined and quantified by Gentilini *et al.* (2015),
who showed that SEM burden increases exponentially with age [@Gentilini2015].
Subsequent work [@Corsaro2023; @Spada2020] extended the framework to disease
association studies, introducing the concept of *lesions* (genomic clusters of
co-occurring SEMs) and *delta metrics* for population-level variability
summaries.

**semseeker** implements this methodology and wraps it into a fully
reproducible R workflow, from raw beta-value matrix to association analysis and
pathway enrichment [@Corsaro2025thesis; @Corsaro2023].

### What semseeker computes

Starting from a normalised beta-value (or M-value) matrix, semseeker:

1. Derives **signal thresholds** (IQR-based) per probe from the reference
   population.
2. Identifies **hypomethylated** and **hypermethylated** SEMs at
   single-probe resolution.
3. Aggregates adjacent SEMs into **lesions** (genomic clusters of
   co-occurring mutations).
4. Computes population-level **delta metrics** (ΔBeta, ΔIV, ΔEV, ΔRV, ΔRP)
   that summarise stochastic epigenetic variability per genomic region.
5. Optionally analyses the **raw methylation signal** directly (mean and range
   per region) for differential methylation without mutation calling.

Supported input platforms: Illumina EPIC (850k), 450k, and 27k arrays
(detected automatically from probe IDs).

<hr>

## Installation

```{r install, eval=FALSE}
# Install remotes if needed
install.packages("remotes")

# Install SEMseeker from GitHub
remotes::install_github("drake69/SEMseeker")
```

> **Note:** semseeker requires the `polars` package from R-multiverse.
> It is installed automatically as a dependency. If you get an installation
> error, try:
> ```r
> Sys.setenv(NOT_CRAN = "true")
> install.packages("polars", repos = "https://community.r-multiverse.org")
> ```

### System dependencies

**macOS — XQuartz.** R 4.6's base `tcltk` package is linked against
`/opt/X11/lib/libX11.6.dylib` at install time. Plain SEMseeker usage
(`library(SEMseeker)`, `semseeker(...)`, `association_analysis(...)`)
does not load tcltk. However, some Bioconductor optional features
that transitively pull `minfi` or `GEOquery` (e.g. running the
annotation-concordance checks) load tcltk and segfault if XQuartz is
not installed. Install XQuartz once via:

```sh
brew install --cask xquartz
```

or download from <https://www.xquartz.org/>. You may need to log out
and back in for the path to be picked up. Linux and Windows users do
not need XQuartz.

**Linux — libuv.** Building from source on Linux pulls `fs` which
needs `libuv` ≥ 1 at compile time. Install via
`sudo apt-get install libuv1-dev` (Debian/Ubuntu) or
`sudo dnf install libuv-devel` (Fedora/RHEL), or set
`USE_BUNDLED_LIBUV=1` to compile against a bundled copy.

<hr>

## Step 1 — Prepare your data

### Beta-value matrix

semseeker expects a **probe × sample** matrix of beta values (0–1) or M-values.
Probe IDs must be Illumina CpG identifiers (e.g. `cg00000029`).

```{r beta_matrix}
# Example: load a pre-normalised matrix (rows = probes, columns = samples).
# Here we use the small synthetic matrix prepared in the hidden setup
# chunk above; in practice you would readRDS("normalised_beta.rds").

dim(beta_matrix)       # e.g. 866895 × 50 on a real EPIC dataset
beta_matrix[1:3, 1:3]
```

If you are starting from raw IDAT files, normalise first with
[ChAMP](https://bioconductor.org/packages/ChAMP/):

```{r chamap_normalise, eval=FALSE}
library(ChAMP)

myLoad <- champ.load(
  directory  = "path/to/idat/",
  arraytype  = "EPIC"
)

myNorm <- champ.norm(
  beta      = myLoad$beta,
  rgSet     = myLoad$rgSet,
  mset      = myLoad$mset,
  method    = "SWAN",
  arraytype = "EPIC"
)

beta_matrix <- myNorm
```

### Sample sheet

A data frame with at minimum:

| Column | Description |
|---|---|
| `Sample_ID` | Unique sample identifier — must match column names of the beta matrix |
| `Sample_Name` | Human-readable label |
| `Sample_Group` | One of `"Case"`, `"Control"`, `"Reference"` |

```{r sample_sheet}
# Example: load a sample sheet (here we use the synthetic one prepared
# in the hidden setup chunk; in practice you would read.csv2("sample_sheet.csv")).

# Minimal required columns
head(sample_sheet[, c("Sample_ID", "Sample_Name", "Sample_Group")])
```

> **Tip:** If you have no independent reference population, duplicate the
> Control rows and set `Sample_Group = "Reference"` for the duplicates.

<hr>

## Step 2 — Detect epimutations

The main entry point is `semseeker()`. A single call runs the full pipeline:
normalisation → mutation calling → lesion aggregation → delta metric computation.

```{r run_semseeker, eval=FALSE}
library(semseeker)

semseeker(
  input         = beta_matrix,
  sample_sheet  = sample_sheet,
  result_folder = "~/semseeker_results/"
)
```

### Key parameters

| Parameter | Default | Description |
|---|---|---|
| `parallel_strategy` | `"multicore"` | Parallelisation backend (`"sequential"`, `"multicore"`, `"multisession"`) |
| `maxResources` | `90` | % of CPU cores to use |
| `iqrTimes` | `3` | IQR multiplier for aberration threshold |
| `alpha` | `0.05` | Significance threshold |
| `areas` | `c("GENE","CHR","ISLAND","PROBE")` | Genomic aggregation levels |
| `marker` | `c("MUTATIONS","LESIONS","DELTARQ",...)` | Which markers to compute |
| `sex_chromosome_remove` | `TRUE` | Remove sex chromosomes from analysis |

Example — run only probe-level mutations, sequential mode, 2 cores:

```{r run_minimal, eval=FALSE}
semseeker(
  input             = beta_matrix,
  sample_sheet      = sample_sheet,
  result_folder     = "~/semseeker_results/",
  parallel_strategy = "sequential",
  areas             = c("PROBE"),
  marker            = c("MUTATIONS", "LESIONS")
)
```

<hr>

## Step 3 — Understand the output

Results are written under `result_folder/Data/`. Each sample produces:

```
Data/
├── SAMPLEID_MUTATIONS_HYPO.bed       ← probe-level hypomethylated mutations
├── SAMPLEID_MUTATIONS_HYPER.bed      ← probe-level hypermethylated mutations
├── SAMPLEID_LESIONS_HYPO.bed         ← genomic lesion clusters (hypo)
├── SAMPLEID_LESIONS_HYPER.bed        ← genomic lesion clusters (hyper)
├── SAMPLEID_SIGNAL_MEAN.PROBE.bedgraph
└── ...
```

Population-level pivot tables aggregate all samples per marker:

```
Data/
├── SIGNAL_MEAN_PROBE_WHOLE.parquet   ← mean signal per probe, all samples
├── MUTATIONS_HYPO_PROBE_WHOLE.parquet
├── DELTARP_GENE_WHOLE.parquet        ← ΔRP per gene, all samples
└── ...
```

### Reading results

```{r read_results, eval=FALSE}
library(polars)

# Read the DELTARP pivot table
deltarp <- as.data.frame(
  pl$read_parquet("~/semseeker_results/Data/DELTARP_GENE_WHOLE.parquet")
)

head(deltarp[, 1:6])
```

<hr>

## Step 4 — Delta metrics explained

semseeker computes five delta metrics per genomic region, each capturing a
different aspect of epigenetic variability:

| Metric | Full name | Captures |
|---|---|---|
| **ΔBeta** | Delta Beta | Mean methylation shift vs reference |
| **ΔIV** | Delta Intra-sample Variability | Within-sample probe dispersion |
| **ΔEV** | Delta Extra-sample Variability | Between-sample variance |
| **ΔRV** | Delta Relative Variability | Normalised inter-sample spread |
| **ΔRP** | Delta Ranked Product | Combined ranked signal — most sensitive |

**ΔRP** (Delta Ranked Product) is the recommended metric for downstream
analysis: it combines mutation burden, lesion frequency and signal variability
into a single ranked score per region, making it robust to outliers and
suitable for association testing.

```{r deltarp_example, eval=FALSE}
# Load ΔRP at gene level
deltarp_gene <- as.data.frame(
  pl$read_parquet("~/semseeker_results/Data/DELTARP_GENE_WHOLE.parquet")
)

# Top regions by mean ΔRP across cases
case_cols  <- grep("^CASE", names(deltarp_gene), value = TRUE)
deltarp_gene$mean_deltarp <- rowMeans(deltarp_gene[, case_cols], na.rm = TRUE)

head(deltarp_gene[order(-deltarp_gene$mean_deltarp),
                  c("CHR", "START", "END", "mean_deltarp")], 10)
```

<hr>

<hr>

## Step 5 — Differential signal analysis

Beyond mutation calling, semseeker also tracks the **raw methylation signal**
per region. Two signal markers are computed automatically:

| Marker | File pattern | Description |
|---|---|---|
| `SIGNAL_MEAN` | `SIGNAL_MEAN_*.parquet` | Per-region mean beta value per sample |
| `SIGNAL_RANGE` | `SIGNAL_RANGE_*.parquet` | Per-region signal spread (max − min) per sample |

These pivot tables have exactly the same structure as the mutation/delta tables
and can be passed directly to `association_analysis()` — allowing you to test
whether mean methylation (or intra-region spread) differs by phenotype, without
any mutation-calling step.

```{r signal_analysis, eval=FALSE}
# Differential mean-signal testing is just another `inference_details` spec:
# the dependent term references the SIGNAL_MEAN marker, the predictor is the
# phenotype, and `depth_analysis = 3` runs it per genomic area (e.g. GENE).
# association_analysis() reads the matching SIGNAL_MEAN pivot table internally —
# no manual parquet loading required.
inference_details <- data.frame(
  formula          = "SIGNAL_MEAN ~ Sample_Group",
  family           = "wilcoxon",   # two independent groups, non-parametric
  transformation_y = "none",
  depth_analysis   = 3,            # per genomic area (region-level / EWAS-style)
  filter_p_value   = TRUE,
  stringsAsFactors = FALSE
)

association_analysis(
  inference_details = inference_details,
  result_folder     = "~/semseeker_results/",
  parallel_strategy = "multicore",
  maxResources      = 90
)
```

This is particularly useful when you want to detect **differential methylation**
at the region level (classic EWAS-style analysis) alongside the epimutation
burden analysis, using a single unified framework.

<hr>

## Subset rationale & full-pipeline runnable

The runnable demo above uses a **subset** of probes (only those around the
BWS-related imprinting DMRs `KCNQ1OT1:TSS-DMR` and `H19_IGF2:IG-DMR`) so the
vignette builds in seconds during CRAN/Bioconductor checks. This is **not**
representative of a real-world workflow: a production analysis would normalise
IDAT files for an entire cohort, correct for biological confounders (age,
cell composition for whole-blood samples), and then run semseeker on the
full beta matrix.

The block below — kept as commented `eval = FALSE` reference — shows the
**complete end-to-end pipeline** as you would run it on a real GEO dataset
such as Beckwith-Wiedemann GSE95486. Uncomment and adapt to your study.

```{r full_pipeline_recipe, eval=FALSE}
# ─── Full pipeline from raw GEO data (~30-60 min on a laptop) ──────────────

## Step 0 — Download IDAT files from GEO (e.g. BWS GSE95486)
library(GEOquery)
gse <- getGEO("GSE95486", GSEMatrix = FALSE, destdir = "./raw")
GEOquery::getGEOSuppFiles("GSE95486", baseDir = "./raw", makeDirectory = TRUE)
# (then extract the IDATs into ./raw/GSE95486/IDATs/)

## Step 1 — Load & SWAN-normalise IDAT files via ChAMP
library(ChAMP)
myLoad <- champ.load(
  directory      = "./raw/GSE95486/IDATs",
  arraytype      = "EPIC",
  filterDetP     = TRUE,
  filterSNPs     = TRUE,
  filterMultiHit = TRUE,
  filterXY       = TRUE
)
myNorm <- champ.norm(
  beta       = myLoad$beta,
  rgSet      = myLoad$rgSet,
  mset       = myLoad$mset,
  method     = "SWAN",
  arraytype  = "EPIC",
  cores      = parallel::detectCores() - 1L
)

## Step 2 — Age correction
##   Confounding by chronological age is the strongest known driver of SEM
##   burden (Gentilini et al., 2015 — SEMs accumulate exponentially with
##   age). Always regress age out before disease-vs-control comparison.
library(limma)
age_residuals <- removeBatchEffect(
  myNorm,
  covariates = sample_sheet$Age   # numeric, years
)

## Step 3 — Cell composition deconvolution (whole-blood samples ONLY)
##   In peripheral whole blood, methylation is strongly confounded by the
##   relative proportions of B cells, CD4/CD8 T, monocytes, NK and granulocytes.
##   For tissues other than whole blood, skip this step.
library(FlowSorted.Blood.EPIC)
cells <- estimateCellCounts2(
  myLoad$rgSet,
  compositeCellType = "Blood",
  referencePlatform = "IlluminaHumanMethylationEPIC"
)
# Regress out the cell-fraction covariates from the (age-residualised) matrix
adjusted <- removeBatchEffect(
  age_residuals,
  covariates = cells$counts
)

## Step 4 — Run SEMseeker on the fully preprocessed matrix
SEMseeker::semseeker(
  input         = adjusted,
  sample_sheet  = sample_sheet,
  result_folder = "./bws_full_results/"
)
# ──────────────────────────────────────────────────────────────────────────
```

**Pre-processing checklist** before running `semseeker()`:

- [ ] Cohort normalisation (SWAN / BMIQ / Functional Normalisation)
- [ ] Detection-p filter, multi-hit/SNP/cross-reactive probe removal
- [ ] **Age** residualisation (mandatory if cohort age range > a few years)
- [ ] **Cell composition** deconvolution (whole-blood only — `FlowSorted.Blood.EPIC`)
- [ ] Batch / processing-date correction (if known)
- [ ] Sample QC pass (detection rate, bead count, predicted-sex check)

semseeker assumes upstream preprocessing has already been performed. Skipping
these steps inflates the false-positive SEM rate — particularly in older
cohorts and whole-blood studies.

<hr>

## References

<!-- automatically generated from getting-started.bib -->

<hr>

## Next steps

- **Association analysis** (all statistical models): see the
  [Association analysis vignette](association-analysis.html).
- **Pathway and enrichment analysis**: see the
  [Pathway analysis vignette](pathway-analysis.html).
- **Full function reference**: see the [Reference](../reference/index.html) page.

<hr>

## Session info

```{r session_info}
sessionInfo()
```
