---
title: "Pathway and enrichment analysis"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: pathway-analysis.bib
vignette: >
  %\VignetteIndexEntry{Pathway and enrichment analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

After `association_analysis()` has identified genomic regions whose SEM
metrics (ΔRP, mutation counts, …) are statistically associated with a
phenotype, the next step is to ask: **which biological pathways and molecular
networks are enriched among those regions?**

semseeker provides six pathway/enrichment functions that all accept the output
of `association_analysis()` as input and run the enrichment for every
combination of genomic area, marker type and sample group that was analysed:

| Function | Backend | Type of analysis |
|---|---|---|
| `pathway_WebGestalt()` | WebGestaltR [@WebGestaltR2023] | Gene Ontology ORA / GSEA |
| `pathway_STRINGdb()` | STRINGdb [@STRINGdb2023] | Protein-protein interaction network |
| `pathway_pathfindR()` | pathfindR [@pathfindR2019] | Active subnetwork enrichment |
| `pathway_ctdR()` | ctdR / CTD [@CTD2023] | Chemical-gene-disease associations |
| `pathway_Phenolyzer_WebGestalt()` | Phenolyzer + WebGestaltR | Disease-prioritised ORA / GSEA |
| `pathway_Phenolyzer_STRINGdb()` | Phenolyzer + STRINGdb | Disease-prioritised network |

All functions are **session-aware**: they read the active semseeker session
(initialised by `association_analysis()`) and therefore do **not** require
passing `result_folder` again — they pick it up automatically via
`get_session_info()`.

<hr>

## Workflow

```
semseeker()               →  epimutation detection + delta metrics
      ↓
association_analysis()    →  statistical associations (CSV in Inference/)
      ↓
pathway_*()               →  enriched pathways / networks (CSV in Pathway/)
      ↓
pathway_cross_studies_compare()       →  cross-cohort Venn diagrams
pathway_cross_subsamples_overlaps()   →  stability across subsamples
```

The `inference_details` object passed to the pathway functions is the **same**
data frame used in `association_analysis()`. It tells the pathway functions
which statistical results to use as the gene list for enrichment.

<hr>

## Installing optional backends

Each pathway function requires a separate optional package. Install only what
you need:

```{r install, eval=FALSE}
# WebGestaltR — ORA / GSEA
install.packages("WebGestaltR")

# STRINGdb — protein network (Bioconductor)
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("STRINGdb")

# pathfindR — active subnetwork enrichment
install.packages("pathfindR")

# ctdR — chemical-disease associations (GitHub)
remotes::install_github("drake69/ctdR")
```

> All backends are in `Suggests` — semseeker installs and runs without them.
> Each function checks for its backend with `requireNamespace()` and logs
> an informative ERROR if it is missing.

<hr>

## 1. `pathway_WebGestalt()` — Gene Ontology ORA and GSEA

WebGestalt [@WebGestaltR2017; @WebGestaltR2023] runs Over-Representation
Analysis (ORA) [@ORA-review] or Gene Set Enrichment Analysis (GSEA)
[@GSEA2005] against Gene Ontology [@GO2000], KEGG [@KEGG2000], Reactome
[@Reactome2020], and other databases.

```{r webgestalt, eval=FALSE}
library(semseeker)

# Run after association_analysis() — session is already active
inference_details <- data.frame(
  formula               = "DELTARP ~ age",
  family                = "spearman",
  transformation_y      = "none",
  depth_analysis        = 3,
  filter_p_value        = TRUE,
  stringsAsFactors      = FALSE
)

association_analysis(
  inference_details = inference_details,
  result_folder     = "~/semseeker_results/",
  parallel_strategy = "multicore"
)

# Gene Ontology ORA on Biological Process and Molecular Function
pathway_WebGestalt(
  study          = "my_study",
  types          = c("BP", "MF"),       # GO namespaces: BP, MF, CC
  enrich_methods = c("ORA"),            # "ORA" or "GSEA"
  pvalue_column  = "PVALUE_ADJ_ALL_BH", # which association p-value to use
  inference_detail = inference_details[1, ],
  significance   = TRUE                 # only use significant genes
)

# GSEA on KEGG pathways
pathway_WebGestalt(
  study          = "my_study",
  types          = c("KEGG"),
  enrich_methods = c("GSEA"),
  pvalue_column  = "PVALUE_ADJ_ALL_BH",
  inference_detail = inference_details[1, ],
  significance   = TRUE
)
```

### Parameters

| Parameter | Default | Description |
|---|---|---|
| `study` | — | Study label, used in output file names |
| `types` | `c("BP","MF")` | GO namespaces or databases: `"BP"`, `"MF"`, `"CC"`, `"KEGG"`, `"Reactome"` |
| `enrich_methods` | `c("ORA")` | `"ORA"` (over-representation) or `"GSEA"` (ranked) |
| `pvalue_column` | `"PVALUE_ADJ_ALL_BH"` | Column from association results used to select / rank genes |
| `inference_detail` | — | One row from `inference_details` |
| `significance` | — | `TRUE` = use only FDR-significant regions as gene list |
| `adjust_per_area` | `FALSE` | Re-adjust p-values within each genomic area |
| `adjust_globally` | `FALSE` | Re-adjust p-values globally across all areas |
| `adjustment_method` | `"BH"` | Multiple testing correction method |

### Output

Results land in `result_folder/Pathway/WebGestalt/` with one CSV per
analysis key. After `enrichment_analysy_add_category()` normalisation, each
result has a unified `SS_CATEGORY` column:

| `SS_CATEGORY` | Source |
|---|---|
| `GO-BP` | Gene Ontology Biological Process |
| `GO-MF` | Gene Ontology Molecular Function |
| `GO-CC` | Gene Ontology Cellular Component |
| `KEGG-ORA` | KEGG via ORA |

<hr>

## 2. `pathway_STRINGdb()` — Protein-protein interaction network

STRINGdb [@STRINGdb2019; @STRINGdb2023] maps the significant genes onto the
STRING protein-protein interaction network and identifies enriched functional
categories, GO terms, KEGG pathways and tissue expressions.

```{r stringdb, eval=FALSE}
pathway_STRINGdb(
  study              = "my_study",
  inference_details  = inference_details,
  pvalue_column      = "PVALUE_ADJ_ALL_BH",
  significance       = TRUE,
  stringDBVersion    = "12.0",           # STRING database version
  statistic_parameter = ""               # "" = presence/absence; column name = use score
)
```

### Parameters

| Parameter | Default | Description |
|---|---|---|
| `study` | — | Study label |
| `inference_details` | — | Full `inference_details` data frame (multi-row supported) |
| `pvalue_column` | `"PVALUE_ADJ_ALL_BH"` | Association p-value column |
| `significance` | `TRUE` | Use only significant regions |
| `stringDBVersion` | `"12.0"` | STRING database version (`"11.5"`, `"12.0"`) |
| `statistic_parameter` | `""` | `""` = unweighted; column name = use that value as interaction weight |
| `adjust_per_area` / `adjust_globally` | `FALSE` | Additional p-value correction |

### Network score threshold

The default score threshold is **200** (medium confidence). Increase to 400
or 700 for higher confidence interactions:

```{r stringdb_score, eval=FALSE}
# Higher confidence network (score >= 700)
# Modify via the STRINGdb$new() call inside the function — currently fixed at 200
# Use statistic_parameter to weight genes by their association score:
pathway_STRINGdb(
  study               = "my_study",
  inference_details   = inference_details,
  statistic_parameter = "SCORE",   # weight by Spearman rho or regression coefficient
  significance        = TRUE
)
```

### Output categories

| `SS_CATEGORY` | Description |
|---|---|
| `GO-BP` | GO Biological Process |
| `GO-MF` | GO Molecular Function |
| `GO-CC` | GO Cellular Component |
| `KEGG-ORA` | KEGG pathway enrichment |

<hr>

## 3. `pathway_pathfindR()` — Active subnetwork enrichment

pathfindR [@pathfindR2019] identifies enriched pathways by searching for
**active subnetworks** in a protein interaction network (PIN) rather than
simple gene list overlap. This approach captures pathway activity even when
only a subset of pathway genes pass the significance threshold.

```{r pathfindr, eval=FALSE}
pathway_pathfindR(
  study              = "my_study",
  path_dbs           = c("KEGG", "Reactome", "GO-BP", "GO-MF", "GO-CC"),
  iterations         = 20,              # number of active subnetwork iterations
  inference_details  = inference_details,
  pvalue_column      = "PVALUE_ADJ_ALL_BH",
  significance       = TRUE,
  statistic_parameter = ""
)
```

### Parameters

| Parameter | Default | Description |
|---|---|---|
| `study` | — | Study label |
| `path_dbs` | — | Databases: `"KEGG"`, `"Reactome"`, `"GO-BP"`, `"GO-MF"`, `"GO-CC"` |
| `iterations` | `20` | Number of active subnetwork search iterations — higher = more stable but slower |
| `inference_details` | — | Full `inference_details` data frame |
| `pvalue_column` | `"PVALUE_ADJ_ALL_BH"` | Association p-value column |
| `significance` | `TRUE` | Filter to significant regions |
| `statistic_parameter` | `""` | Optional: column to use as gene-level score |

### Output categories

| `SS_CATEGORY` | Description |
|---|---|
| `GO-BP` | GO Biological Process (active subnetwork) |
| `GO-MF` | GO Molecular Function (active subnetwork) |
| `KEGG-NTA` | KEGG via network topology (NTA) |
| `REACTOME-NTA` | Reactome via network topology |

### Installation note

pathfindR requires `ggkegg` (Bioconductor). Install the full chain:

```{r pathfindr_install, eval=FALSE}
BiocManager::install("ggkegg")
install.packages("pathfindR")
```

<hr>

## 4. `pathway_ctdR()` — Chemical-gene-disease associations

`pathway_ctdR()` uses the **Comparative Toxicogenomics Database** (CTD)
[@CTD2023] via the `ctdR` package to find chemical compounds and diseases
associated with the significant genes. This is particularly useful for
epigenetic studies involving environmental exposures.

```{r ctdr, eval=FALSE}
pathway_ctdR(
  study              = "my_study",
  inference_details  = inference_details,
  pvalue_column      = "PVALUE_ADJ_ALL_BH",
  significance       = TRUE,
  statistic_parameter = ""
)
```

### Parameters

| Parameter | Default | Description |
|---|---|---|
| `study` | — | Study label |
| `inference_details` | — | Full `inference_details` data frame |
| `pvalue_column` | `"PVALUE_ADJ_ALL_BH"` | Association p-value column |
| `significance` | `TRUE` | Use only significant regions |
| `statistic_parameter` | `""` | Optional signal weight column |

### Output

Results are normalised to `SS_CATEGORY = "CHEMICAL"`, linking significant
epigenetic regions to chemical exposures and associated diseases.

<hr>

## 5 & 6. Phenolyzer variants — disease-prioritised enrichment

The Phenolyzer variants (`pathway_Phenolyzer_WebGestalt()` and
`pathway_Phenolyzer_STRINGdb()`) first **prioritise genes** using the
Phenolyzer disease-gene scoring system and then run enrichment on the
prioritised list.

Use these when you have a specific disease hypothesis and want to focus the
enrichment on genes already known to be relevant to that disease:

```{r phenolyzer, eval=FALSE}
# Phenolyzer + WebGestalt
pathway_Phenolyzer_WebGestalt(
  study            = "my_study",
  disease          = "breast cancer",   # disease name for Phenolyzer prioritisation
  types            = c("BP", "MF"),
  enrich_methods   = c("ORA"),
  pvalue_column    = "PVALUE_ADJ_ALL_BH",
  inference_detail = inference_details[1, ],
  significance     = TRUE
)

# Phenolyzer + STRINGdb
pathway_Phenolyzer_STRINGdb(
  study               = "my_study",
  disease             = "breast cancer",
  inference_detail    = inference_details,
  pvalue_column       = "PVALUE_ADJ_ALL_BH",
  significance        = TRUE,
  statistic_parameter = ""
)
```

### Additional parameter: `disease`

| Parameter | Description |
|---|---|
| `disease` | Free-text disease name passed to Phenolyzer for gene prioritisation (e.g. `"lung cancer"`, `"Alzheimer"`, `"TCDD exposure"`) |

<hr>

## Running all backends in one script

```{r all_backends, eval=FALSE}
library(semseeker)

result_folder <- "~/semseeker_results/"

# 1. SEM detection
semseeker(sample_sheet, beta_matrix, result_folder)

# 2. Association analysis
inference_details <- data.frame(
  formula          = "DELTARP ~ case_control",
  family           = "wilcoxon",
  transformation_y = "none",
  depth_analysis   = 3,
  filter_p_value   = TRUE,
  stringsAsFactors = FALSE
)
association_analysis(inference_details, result_folder)

# 3. Pathway enrichment — run all backends
pathway_WebGestalt(
  study="study1", types=c("BP","MF","KEGG"), enrich_methods=c("ORA"),
  pvalue_column="PVALUE_ADJ_ALL_BH",
  inference_detail=inference_details[1,], significance=TRUE)

pathway_STRINGdb(
  study="study1", inference_details=inference_details,
  pvalue_column="PVALUE_ADJ_ALL_BH", significance=TRUE,
  stringDBVersion="12.0")

pathway_pathfindR(
  study="study1", path_dbs=c("KEGG","Reactome","GO-BP"),
  iterations=20, inference_details=inference_details,
  pvalue_column="PVALUE_ADJ_ALL_BH", significance=TRUE)

pathway_ctdR(
  study="study1", inference_details=inference_details,
  pvalue_column="PVALUE_ADJ_ALL_BH", significance=TRUE)
```

<hr>

## Output structure

All results are written under `result_folder/Pathway/`:

```
Pathway/
├── WebGestalt/
│   └── <samples_filter>/<inference_filter>/
│       └── <area>_<marker>_<figure>_<test>_<variable>.csv
├── STRINGdb/
│   └── ...
├── pathfindR/
│   └── ...
├── ctdR/
│   └── ...
├── Phenolyzer_WebGestalt/
│   └── ...
└── Phenolyzer_STRINGdb/
    └── ...
```

Each CSV is processed by `enrichment_analysy_add_category()` which:
- Normalises category labels to `SS_CATEGORY` (consistent across backends)
- Adds a `PHENOTYPE` column (`TRUE` if the term matches the study name)
- Applies multiple testing correction (BH or q-value)
- Sorts by adjusted p-value

### Common output columns

| Column | Description |
|---|---|
| `SS_CATEGORY` | Normalised category (`GO-BP`, `KEGG-ORA`, `KEGG-NTA`, `REACTOME-NTA`, `CHEMICAL`, …) |
| `ID` | Pathway / term identifier |
| `Description` | Human-readable term name |
| `SCORE` | Enrichment score (ORA: enrichment ratio; GSEA: NES; STRINGdb: FDR) |
| `PVALUE_ADJ_ALL_BH` | BH-adjusted p-value across all areas |
| `PHENOTYPE` | `TRUE` if term description matches study label |
| `SS_RANK` | Rank within `SS_CATEGORY` |

<hr>

## Cross-study and cross-subsample comparison

### `pathway_cross_studies_compare()`

Compares enriched pathways across two or more independent cohorts and produces
Venn diagrams [@VennDiagram] of overlapping terms:

```{r cross_studies, eval=FALSE}
pathway_cross_studies_compare(
  result_folder = "~/meta_results/",
  # studies A and B must already have Pathway/ results
)
```

### `pathway_cross_subsamples_overlaps()`

Tests the **stability** of pathway enrichment under random subsampling of the
cohort. Run this to assess how robust the enriched pathways are to sample
variability:

```{r cross_subsamples, eval=FALSE}
pathway_cross_subsamples_overlaps(
  inference_details  = inference_details,
  result_folder      = "~/semseeker_results/",
  pathway_package    = "WebGestalt",            # "" = compare all backends
  significance       = TRUE,
  pathways_sql_selection = "",                  # SQL to filter specific pathways
  old_label_samples_sql_condition    = "",
  new_label_samples_sql_condition    = "",
  association_pvalue_column          = "PVALUE_ADJ_ALL_BH"
)
```

<hr>

## Choosing the right backend

| Scenario | Recommended backend |
|---|---|
| First exploratory run, GO + KEGG | `pathway_WebGestalt()` with ORA |
| Small gene lists (< 20 genes) | `pathway_WebGestalt()` with ORA |
| Ranked gene lists, pathway activity | `pathway_WebGestalt()` with GSEA |
| Protein interaction context | `pathway_STRINGdb()` |
| Partial pathway activation (sparse hits) | `pathway_pathfindR()` |
| Environmental/chemical exposure study | `pathway_ctdR()` |
| Known disease hypothesis | `pathway_Phenolyzer_WebGestalt()` or `pathway_Phenolyzer_STRINGdb()` |
| Replication across cohorts | `pathway_cross_studies_compare()` |
| Bootstrap stability check | `pathway_cross_subsamples_overlaps()` |

## Session info

```{r session_info}
sessionInfo()
```

<hr>

## References
